---
title: An Analysis of Sexual Dimorphism in Gentoo Penguins (Pygoscelis papua)
author: 
- name: Duygu Nur Arabaci
  affiliation: Department of Bioinformatics, University of Potsdam, Germany
- name: Mekontso Tepeu Florentin Jose
  affiliation: Department of Bioinformatics, University of Potsdam, Germany
bibliography: [penguins_refs.bib, packages.bib]

nocite: '@*'

output:
  bookdown::html_document2:
    theme: yeti
    toc: true
    toc_depth: 2
  number_sections: TRUE
  fig_caption: TRUE



---

```{r include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message = FALSE, tidy = TRUE)
```

```{r label=loading,results="hide"}
# Packages
packages <- c("palmerpenguins", "ggplot2", "knitr", "tidyverse", "lsr", "bookdown", "vcd")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

source("arabaci2_lipsum.R")
source("sbi.R")

#Create citations for packages used
knitr::write_bib(c(.packages(), packages), "packages.bib")
```


-------------------------------------------------------------

**Abstract**

This research aims at examining discernible variations in physical traits and behaviors between male and female members of the  Gentoo penguins.

-------------------------------------------------------------

<div id="abstract">
</div>
<br>

# Introduction


Sexual dimorphism, denoting observable differences in form between male and female individuals of the same species, encompasses a range of attributes such as size, color, and behavioral tendencies. Initially identified by Rensch, who noted a general trend of larger size in males across species, sexual dimorphism primarily arises from sexual selection rather than survival advantage. This study focuses on Gentoo penguins, belonging to the Pygoscelis genus, within the marine ecosystem of the western Antarctic peninsula, exploring their sexual dimorphism with a specific emphasis on body mass, bill features, and flipper dimensions. Vocal signals play a crucial role in Gentoo penguins' reproductive success, with males utilizing an "Ecstatic Display Call" during courtship, identified as a significant indicator of male quality and predictive of reproductive success (Marks, Brunton, & Rodrigo, 2010).

In line with existing literature, Gorman et al.'s 2014 study emphasizes body mass as a strong predictor of sex among Gentoo and Adelie penguins, with culmen length and depth being influential for different species within the Pygoscelis genus. Notably, Chinstrap penguins exhibit the most apparent sexual size dimorphism compared to Adelie and Gentoo penguins (Gorman, K. B., Williams, T. D., & Fraser, W. R., 2014). Adelie penguins, characterized by their exceptional swimming abilities and an average length of 70-73 cm, exhibit sexual dimorphism with larger males, and their average lifespan is around 16 years (Rafferty, 2020a).

Chinstraps, inhabitants of islands around the Southern Pacific and Antarctic oceans, showcase similar sexual dimorphism trends, with males being larger than females and an average lifespan of around 16 years (Rafferty, 2020a). Gentoo penguins, closely related to Adelie penguins, are notable for their larger size, standing 75-95 cm tall, and weighing between 5 to 8.5 kgs. The world's fastest underwater birds, Gentoo males are larger than females, capable of reaching speeds up to 36 km/h (Rafferty 2020c).

All three species of the Pygoscelus genus—Adelie, Chinstrap, and Gentoo—display monogamous behavior, sharing incubation duties during the 5-week period it takes for eggs to hatch. Their reproductive cycles differ slightly, with Gentoo penguins laying eggs in October-November and hatching in February. This project establishes four hypotheses related to Gentoo penguins' sexual dimorphism, examining body mass, bill dimensions, and flipper length, and their potential correlations with reproductive success.

In this paper, our aim is to elucidate the sexual dimorphism of Gentoo penguins. Thus, we will investigate the hypotheses that in Gentoo penguins: (1) body mass is associated with sex , (2) bill length is associated with sex, (3) flipper length is associated with sex.

# Data

palmerpenguins package


## General information about Palmer Penguins data
```{r}
penguins=as.data.frame(penguins)
#number of penguins
total_pengs=length(penguins$species)
pengs = penguins %>% count(species)
adelies = pengs[1,2]
chinstraps = pengs[2,2]
gentoos = pengs[3,2]

tab = table(penguins$species, penguins$island)
```

```{r tab}
knitr::kable(tab,caption = "The number of penguin species on the three islands.", "pipe", align = "ccc")
```

# Statistical Analysis

In this report, we chose α=0.05 as significance threshold. Data distribution normality was investigated with histograms and Shapiro-Wilkes normality test. Kurtosis and skewness of distributions of penguin features were examined across sexes. Due to non-normal distribution of data, Wilcoxon test was chosen to investigate the difference of means between males and females.

We categorized each feature into three categories and compared them in 3x2 contingency tables. To investigate dependence between sex and body mass, bill length, bill depth and flipper length in categorical format, Fisher's Exact test was chosen over χ2 because some frequencies in the data were ≤ 5. To elucidate effect size, Cramer's V values were determined. Used in 3x2 tables, Cramer's V values for effect size (ES) <0.10 were considered weakly associated, 0.1 < ES ≤ 0.5 was considered moderately associated, and ES > 0.5 was considered strongly associated. For association strength, Pearson residuals were calculated.



# Results

## General statistics of the Gentoo sample

```{r}
#On Gentoos
Gentoo=penguins[penguins$species=="Gentoo",]

male=Gentoo[Gentoo$sex=="male",]
female=Gentoo[Gentoo$sex=="female",]
#Gentoo number of males and females

```
```{r}
mean_bill_depth = as.double(round(mean(Gentoo$bill_depth_mm, na.rm = TRUE)))
mean_bill_length = as.double(round(mean(Gentoo$bill_length_mm, na.rm = TRUE)))
mean_body_mass = as.double(round(mean(Gentoo$body_mass_g, na.rm = TRUE)))
mean_flipper_length = as.double(round(mean(Gentoo$flipper_length_mm, na.rm = TRUE)))
means_gentoo = cbind(mean_body_mass,mean_bill_length ,mean_bill_depth, mean_flipper_length)

means_sexes = as.table(rbind(c(round(mean(Gentoo$body_mass_g, na.rm=TRUE), digit=2), 
                            round(mean(Gentoo$bill_length_mm, na.rm=TRUE), digit=2), 
                            round(mean(Gentoo$bill_depth_mm, na.rm=TRUE), digit=2), 
                            round(mean(Gentoo$flipper_length_mm, na.rm=TRUE), digit=2))
                            ,c(round(mean(male$body_mass_g, na.rm=TRUE), digit=2), 
                            round(mean(male$bill_length_mm, na.rm=TRUE), digit=2), 
                            round(mean(male$bill_depth_mm, na.rm=TRUE), digit=2), 
                            round(mean(male$flipper_length_mm, na.rm=TRUE), digit=2)), 
                            c(round(mean(female$body_mass_g, na.rm=TRUE), digit=2), 
                            round(mean(female$bill_length_mm, na.rm=TRUE, digit=2)), 
                            round(mean(female$bill_depth_mm, na.rm=TRUE), digit=2), 
                            round(mean(female$flipper_length_mm, na.rm=TRUE), digit=2))))

rownames(means_sexes)=c("Combined","Male", "Female")
colnames(means_sexes)=c("Body mass (in g)", "Bill length (in mm)", "Bill depth (in mm)", "Flipper length (in mm)")

knitr::kable(means_sexes)

```

The total number of penguins was `r total_pengs`. A total of `r adelies`, `r gentoos`, `r chinstraps` penguins of species Adelie, Gentoo and Chinstraps were included respectively.


```{r}
#Defining color variables here so that they are consistent and can be changed easily later on
color_f = "hotpink2"
color_m  = "skyblue"
```




For inferential statistics, we first consider distribution of data (normal or non-normal) and number of categories. (paired or unpaired?). 

```{r}
#check if data is normally distributed
par(mfrow = c(2,4), mai = c(0.6, 0.2, 0.5, 0.2))
hist(male$body_mass_g, xlab = "Male Body Mass (g)", main = "Male Body Mass", col=color_m, border="black")
hist(female$body_mass_g, xlab = "Female Body Mass (g)", main = "Female Body Mass", col=color_f, border="black")

hist(male$bill_length_mm, xlab = "Male Bill Length (mm)", main = "Male Bill Length", col=color_m, border="black")
hist(female$bill_length_mm, xlab = "Female Bill Length (mm)", main = "Female Bill Length", col=color_f, border="black")

hist(male$bill_depth_mm, xlab = "Male Bill Depth (mm)", main = "Male Bill Depth", col=color_m, border="black")
hist(female$bill_depth_mm, xlab = "Female Bill Depth (mm)", main = "Female Bill Depth", col=color_f, border="black")

hist(male$flipper_length_mm, xlab = "Male Flipper Length (mm)", main = "Male Flipper Length", col=color_m, border="black")
hist(female$flipper_length_mm, xlab = "Female Flipper Length (mm)", main = "Female Flipper Length", col=color_f, border="black")
```

```{r}
k.mbl = sbi$kurtosis(male$bill_length_mm)
k.fbl = sbi$kurtosis(female$bill_length_mm)

s.mbl = sbi$skewness(male$bill_length_mm)
s.fbl = sbi$skewness(female$bill_length_mm)
```

```{r}
k.mbd = sbi$kurtosis(male$bill_depth_mm)
k.fbd = sbi$kurtosis(female$bill_depth_mm)

s.mbd = sbi$skewness(male$bill_depth_mm)
s.fbd = sbi$skewness(female$bill_depth_mm)
```

```{r}
k.mfl = sbi$kurtosis(male$flipper_length_mm)
k.ffl = sbi$kurtosis(female$flipper_length_mm)

s.mfl = sbi$skewness(male$flipper_length_mm)
s.ffl = sbi$skewness(female$flipper_length_mm)
```

```{r}
k.mbm = sbi$kurtosis(male$body_mass_g)
k.fbm = sbi$kurtosis(female$body_mass_g)

s.mbm = sbi$skewness(male$body_mass_g)
s.fbm = sbi$skewness(female$body_mass_g)
```

```{r}
kurt_skew_summary = as.table(cbind(rbind(c(k.mbm,k.fbm), c(s.mbm, s.fbm)),rbind(c(k.mbd,k.fbd), c(s.mbd, s.fbd)), rbind(c(k.mbl,k.fbl), c(s.mbl, s.fbl)),  rbind(c(k.mfl,k.ffl), c(s.mfl, s.ffl))))
colnames(kurt_skew_summary) = c("Male Body Mass","Female Body Mass","Male Bill Depth","Female Bill Depth",  "Male Bill Length","Female Bill Length", "Male Flipper Length","Female Flipper Length")
rownames(kurt_skew_summary) = c("Kurtosis","Skewness")

knitr::kable(
  kurt_skew_summary, 
  caption = "Summary of Gentoo Features Data Distribution Description", 
  "pipe", 
  align = "ccccccccc", 
  digits = 2) 
```


```{r}
#defining the tables 
cWeight=cut(Gentoo$body_mass_g,breaks=quantile(Gentoo$body_mass_g,c(0,1/3,2/3,1),na.rm=TRUE),include.lowest=TRUE)
levels(cWeight)=c("Light","Medium","Heavy")
weight_cat = table(cWeight, Gentoo$sex)
colnames(weight_cat) = c("Female", "Male")
```


```{r}
cDepthBill=cut(Gentoo$bill_depth_mm,breaks=quantile(Gentoo$bill_depth_mm,c(0,1/3,2/3,1),na.rm=TRUE),include.lowest=TRUE)
levels(cDepthBill)=c("Shallow","Medium","Deep")
bdepth_cat = table(cDepthBill, Gentoo$sex)
colnames(bdepth_cat) = c("Female", "Male")
```

```{r}
cLengthBill=cut(Gentoo$bill_length_mm,breaks=quantile(Gentoo$bill_length_mm,c(0,1/3,2/3,1),na.rm=TRUE),include.lowest=TRUE)
levels(cLengthBill)=c("Shallow","Medium","Deep")
blength_cat = table(cLengthBill, Gentoo$sex)
colnames(blength_cat) = c("Female", "Male")
```

```{r}
cLengthFlipper=cut(Gentoo$flipper_length_mm,breaks=quantile(Gentoo$flipper_length_mm,c(0,1/3,2/3,1),na.rm=TRUE),include.lowest=TRUE)
levels(cLengthFlipper)=c("Small","Medium","Large")
flength_cat = table(cLengthFlipper, Gentoo$sex)
colnames(flength_cat) = c("Female", "Male")
```



```{r}
bl_m = shapiro.test(male$bill_length_mm)
bl_f = shapiro.test(female$bill_length_mm)
bd_m = shapiro.test(male$bill_depth_mm)
bd_f = shapiro.test(female$bill_depth_mm)
#(p<0.05 means not normally distributed)
s_ps = round(c(bl_m$p.value, bl_f$p.value, bd_m$p.value, bd_f$p.value), digits = 2)

```
Shapiro-Wilk normality test was performed on male and female bill length and bill depth data. The p values for bill length distributions for males and females respectively were `r s_ps[1]` and `r s_ps[2]` respectively; and the p values for bill depth distributions for males and females were `r s_ps[3]` and `r s_ps[4]`, respectively. 
Shapiro-Wilk normality tests for combined bill depth and bill length have shown p values of `r round(shapiro.test(Gentoo$bill_depth_mm)$p.value, digits=2)` and `r round(shapiro.test(Gentoo$bill_length_mm)$p.value, digits=2)`, both less than 0.05, suggesting non-normal distribution.With p<0.05, null hypothesis that they come from normally distributed data is rejected. Thus, a Wilcox test is appropriate.

```{r}
w_weight.p =round(wilcox.test(weight_cat)$p.value, digits=3)
```
We conducted a Wilcoxon signed rank test to compare the body mass of Gentoo penguins between sexes. The body mass measurements of female Gentoo penguins were significantly smaller than the male Gentoo measurements (p=`r w_weight.p`).

## Association of Body Weight and Sex

```{r}
knitr::kable(weight_cat, caption = "Number of Gentoos in each weight category", "pipe", align = "ccc") 
```

Since some cells in the contingency tables are <5, fisher's exact test was appropriate to determine relation.

```{r}
fisher.body_weight_sex.p = fisher.test(weight_cat)$p.value
```
We conducted a Fisher's Exact Test for Count Data to determine dependence between body mass and sex of Gentoo penguins. The results showed significant dependance, p<0.0001. 


```{r fig.width=7,fig.height=5, fig.cap="Body Mass of Males and Females"}
layout(matrix(c(1,2,3,3), ncol=2, byrow=TRUE), heights=c(4, 1))

with(Gentoo,boxplot(Gentoo$body_mass_g ~ sex,col=c(color_f, color_m), ylab="Body Mass (in g)"))
barplot(t(weight_cat),xlab="Weight Categories", ylab = "Number of Penguins", col=c(color_f, color_m))

par(mai=c(0,0,0,0))
plot.new()
legend(x="center", ncol=2,legend=c("Male","Female"),
       fill=c(color_m, color_f), title="Penguins body weight by sex")
```
```{r}
assoc.gentoo_weight = table(Gentoo$sex, cWeight)
cramersV.gentoo_weight = round(sbi$cramersV(assoc.gentoo_weight), digits=3) #from sbi.R
```

A Cramer's V test was performed to examine the relationship between sex and body weight in Gentoo penguins and determine the effect size. There was a strong influence of the sex on body weight category distribution between all weight classes p<0.05, Cramer's V is `r cramersV.gentoo_weight`.

```{r}
dimnames(assoc.gentoo_weight) = list("Sex"=dimnames(assoc.gentoo_weight)[[1]],
                                     "Penguin Body Mass" = dimnames(assoc.gentoo_weight)[[2]])
assoc(assoc.gentoo_weight, shade=TRUE)

```

## Association of Bill Depth and Sex

```{r}
knitr::kable(bdepth_cat, caption = "Number of Gentoos in each Bill Depth category", "pipe", align = "ccc") 
```

Since some cells in the contingency tables are <5, fisher's exact test was appropriate to determine relation.

```{r}
fisher.bill_depth_sex = fisher.test(bdepth_cat)
```
We conducted a Fisher's Exact Test for Count Data to determine dependence between bill depth and sex of Gentoo penguins. The results showed significant dependance, p<0.0001. 


```{r}
assoc.gentoo_bdepth = table(Gentoo$sex, cDepthBill)
cramersV.gentoo_bdepth = round(sbi$cramersV(assoc.gentoo_bdepth), digits=3) #from sbi.R
```

Cramer's V is `r cramersV.gentoo_bdepth`.

```{r fig.width=7,fig.height=5, fig.cap="Bill Depth of Males and Females"}
layout(matrix(c(1,2,3,3), ncol=2, byrow=TRUE), heights=c(4, 1))

with(Gentoo,boxplot(Gentoo$bill_depth_mm ~ sex,col=c(color_f, color_m), ylab="Bill Depth (in mm)"))
barplot(t(bdepth_cat),xlab="Bill Depth Categories", ylab = "Number of Penguins", col=c(color_f, color_m))

par(mai=c(0,0,0,0))
plot.new()
legend(x="center", ncol=2,legend=c("Male","Female"),
       fill=c(color_m, color_f), title="Penguins bill depth by sex")
```

```{r}
dimnames(assoc.gentoo_bdepth) = list("Sex"=dimnames(assoc.gentoo_bdepth)[[1]],
                                     "Penguin Bill Depth" = dimnames(assoc.gentoo_bdepth)[[2]])
assoc(assoc.gentoo_bdepth, shade=TRUE)

```


## Association of Bill Length and Sex

```{r}
knitr::kable(blength_cat, caption = "Number of Gentoos in each Bill Length category", "pipe", align = "ccc") 
```
Since some cells in the contingency tables are <5, fisher's exact test was appropriate to determine relation.

```{r}
fisher.bill_length_sex = fisher.test(blength_cat)
```
We conducted a Fisher's Exact Test for Count Data to determine dependence between bill length and sex of Gentoo penguins. The results showed significant dependance, p<0.0001. 

```{r fig.width=7,fig.height=5, fig.cap="Bill Length of Males and Females"}
layout(matrix(c(1,2,3,3), ncol=2, byrow=TRUE), heights=c(4, 1))

with(Gentoo,boxplot(Gentoo$bill_length_mm ~ sex,col=c(color_f, color_m), ylab="Bill Length (in mm)"))
barplot(t(blength_cat),xlab="Bill Length Categories", ylab = "Number of Penguins", col=c(color_f, color_m))

par(mai=c(0,0,0,0))
plot.new()
legend(x="center", ncol=2,legend=c("Male","Female"),
       fill=c(color_m, color_f), title="Penguins bill length by sex")
```

```{r}
assoc.gentoo_blen = table(Gentoo$sex, cLengthBill)
cramersV.gentoo_blen = round(sbi$cramersV(assoc.gentoo_blen), digits=3) #from sbi.R
```


Cramer's V is `r cramersV.gentoo_blen`.

```{r}
dimnames(assoc.gentoo_blen) = list("Sex"=dimnames(assoc.gentoo_blen)[[1]],
                                     "Penguin Bill Length" = dimnames(assoc.gentoo_blen)[[2]])
assoc(assoc.gentoo_blen, shade=TRUE)
```
## Flipper Length and Sex

```{r}
knitr::kable(flength_cat, caption = "Number of Gentoos in each Flipper Length category", "pipe", align = "ccc") 
```
Since some cells in the contingency tables are <5, fisher's exact test was appropriate to determine relation.

```{r}
fisher.flipper_length_sex = fisher.test(flength_cat)
```
We conducted a Fisher's Exact Test for Count Data to determine dependence between bill depth and sex of Gentoo penguins. The results showed significant dependance, p<0.0001. 

```{r fig.width=7,fig.height=5, fig.cap="Flipper Length of Males and Females"}
layout(matrix(c(1,2,3,3), ncol=2, byrow=TRUE), heights=c(4, 1))

with(Gentoo,boxplot(Gentoo$flipper_length_mm ~ sex,col=c(color_f, color_m), ylab="Flipper Length (in mm)"))
barplot(t(flength_cat),xlab="Flipper Length Categories", ylab = "Number of Penguins", col=c(color_f, color_m))

par(mai=c(0,0,0,0))
plot.new()
legend(x="center", ncol=2,legend=c("Male","Female"),
       fill=c(color_m, color_f), title="Penguins flipper length by sex")
```

```{r}
assoc.gentoo_flength = table(Gentoo$sex, cLengthFlipper)
cramersV.gentoo_flength = round(sbi$cramersV(assoc.gentoo_flength), digits=3) #from sbi.R
```

Cramer's V is `r cramersV.gentoo_flength`.

```{r}
dimnames(assoc.gentoo_flength) = list("Sex"=dimnames(assoc.gentoo_flength)[[1]],
                                     "Penguin Flipper Length" = dimnames(assoc.gentoo_flength)[[2]])
assoc(assoc.gentoo_flength, shade=TRUE)

```


## Summary 

```{r}
cramersV_list = as.table(c(cramersV.gentoo_weight,cramersV.gentoo_blen, cramersV.gentoo_bdepth, cramersV.gentoo_flength))
rownames(cramersV_list) = c("Body Mass", "Bill Length", "Bill Depth", "Flipper Length")
knitr::kable(cramersV_list, col.names = c("", "Cramer's V Values") , caption = "Association of penguin features with sex","pipe", align = "lccc")
```

```{r fig.width=7,fig.height=5, fig.cap="Paired up association of Gentoo features Summary"}
par(mai = c(0.5,0.5,0.5,0.5))
pairs(~ bill_length_mm+ bill_depth_mm + body_mass_g + flipper_length_mm ,data = Gentoo,col = c(color_m, color_f) )
```

```{r}
```


# Discussion

`r lipsum(2,paragraphs=1)`

# Acknowledgment

`r lipsum(2,paragraphs=1)`

# Bibliography

<div id="refs"></div>

# Appendix

