---
title: Sexual Dimorphism of Penguins
author: 
- name: Duygu Nur Arabaci
  affiliation: Department of Bioinformatics, University of Potsdam, Germany
- name: Mekontso Tepeu Florentin Jose
  affiliation: Department of Bioinformatics, University of Potsdam, Germany
bibliography: [penguins_refs.bib, packages.bib]

nocite: '@*'

output:
  bookdown::html_document2:
  number_sections: TRUE
  fig_caption: TRUE

---

```{r include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message = FALSE, tidy = TRUE)
```

```{r label=loading,results="hide"}
# Packages
packages <- c("palmerpenguins", "ggplot2", "knitr", "tidyverse", "lsr", "bookdown")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

source("arabaci2_lipsum.R")

#Create citations for packages used
knitr::write_bib(c(.packages(), packages), "packages.bib")
```

# Abstract

<div id="abstract">
`r lipsum(2,paragraphs=1, lang="english")`
</div>
<br>

# Introduction

`r lipsum(2,paragraphs=1, lang="english")`

# Data and Methods

## General information about Palmer Penguins data
```{r}
penguins=as.data.frame(penguins)
#number of penguins
total_pengs=length(penguins$species)
pengs = penguins %>% count(species)
adelies = pengs[1,2]
chinstraps = pengs[2,2]
gentoos = pengs[3,2]

tab = table(penguins$species, penguins$island)
```

```{r tab}
knitr::kable(tab,caption = "The number of penguin species on the three islands.")
```

The total number of penguins was `r total_pengs`. A total of `r adelies`, `r gentoos`, `r chinstraps` penguins of species Adelie, Gentoo and Chinstraps were included respectively.

# Statistical Analysis

`r lipsum(2,paragraphs=1)`

# Results

`r lipsum(2,paragraphs=1)`

# Discussion

`r lipsum(2,paragraphs=1)`

# Acknowledgment

`r lipsum(2,paragraphs=1)`

# Bibliography

<div id="refs"></div>

# Appendix

