---
title: Sexual Dimorphism of Penguins
author: Duygu Nur Arabaci, University of Potsdam, Germany; Mekontso Tepeu Florentin Jose, University of Potsdam, Germany
deadline: 2023-12-29
---


```{r}
# Packages
packages <- c("palmerpenguins", "ggplot2", "knitr")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

source("lipsum.R")
```

```{r include=FALSE}
knitr::opts_chunk$set(comment = "###")
```

```{r label=loading,results="hide",echo=FALSE}
### load required packages:
library(knitr)
library(palmerpenguins)
```
**Abstract**
```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```


## Introduction


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```


## Data and Methods


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```

## Statistical Analysis


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```


## Results


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```

## Discussion


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```

## Acknowledgment


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```


## References


```{r lipsum(2,paragraphs=3)}
knitr::opts_chunk$set(comment = "###")
```


## R Examples

Delete the text which follows at the end:

Data loading

```{r label=data}
data(iris)
summary(iris)
```

Table example:

```{r label=table,results="asis"}
knitr::kable(head(iris))
```

Figure example:

```{r label=fig1,fig.width=9,fig.height=6,out.width=800}
with(iris,boxplot(Sepal.Length ~ Species,col=2:4))
```

Inline code example: We have `r nrow(iris)` plants in the dataset.

## Markdown formatting

* list item
* list item

__bold__, __italic__, `typewriter`

> indented

> * indented list
  * indented list

| col1  | col2 | col 3 |
|:-----:|:----:|:-----:|
|   1   |  2   | 3    |
|   4   |  5   | 6    |


Some more text then here ...

## Document processing

This document can be processed to HTML using the packages knitr and markdown:

```
library(knitr);
knitr::knit2html('template.Rmd',output='report.md');
### which should create a file report.html in your current directory
```

If you work on a Unix system We can as well create a bash script and place it into your personal bin folder.

```
#!/usr/bin/env bash
function rmd2html {
  if [ -z $1 ]; then
    echo "Usage: rmd2html RMDFILE"
  else
    echo "library(knitr);knitr::knit2html('$1');" | Rscript -
  fi
}
# delegate arguments to the function
rmd2html "$@"
```

If you have the rmarkdown package and pandoc installed you can do this:

```
library(rmarkdown);
rmarkdown::render('template.Rmd',html_document())
```

My rmarkdown package is currently broken more features, more often problems ...

